Assessing the Accessibility of Protein Tunnels using Motion Planning

### 2017-09-10 paper.tex
reference reorganized
shortening

The idea of the sampling-based motion planning is to randomly sample the configuration space of the robot and to connect the collision-free samples
into a graph structure (roadmap).
A path in the roadmap then corresponds to a motion in the workspace.
%The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
%The sampling-based planners can cope with robots with many degrees of freedom (DOF).
Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
Therefore, they have been used for motion planning of various robotic systems.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
%Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
%In bio-chemistry, the planners have been used to study
to study loop motions~\cite{cortes2004geometric}
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
and for protein folding~\cite{raveh2009rapid,novinskaya2015improving}.
%or for tunnel detection~\cite{vonasek2017tunnel}.

@@

In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
packed conformations can be used.

\begin{figure}[b]
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The results are summarized in Tab.~\ref{tab::main}.
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.
The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).

\begin{table}[ht]
\centering
\caption{\label{tab::main}
    Traversability of the tunnels for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graph.tex
}
\end{table}

@@

The most time-consuming part of the planners is the collision detection. 
Although both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
The reason is that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
it allows the ligands to retract more towards $\qrand$.
On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
%To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
%prevent the usage KD-tree for the nearest-neighbor search.

>>>>

The idea of the sampling-based motion planning is to randomly sample the configuration space of the robot and to connect the collision-free samples
into a graph structure (roadmap).
A path in the roadmap then corresponds to a motion in the workspace.
%The random samples are classified as free or non-free using black-box collision detection which allows the sampling-based methods to find paths for robots of various shapes. 
%The sampling-based planners can cope with robots with many degrees of freedom (DOF).
Two most commonly used sampling-based techniques are Probabilistic Roadmaps (PRM)~\cite{kavrakiForPP} and Rapidly Exploring Random Trees (RRT)~\cite{lavalleRRT}.
The sampling-based planners can cope with many-DOF (Degrees of Freedom) robots of an arbitrary shape.
Therefore, they have been used for motion planning of various robotic systems.
Besides many applications in robotics, the sampling-based planners have been already used in biochemistry as well, 
%Sampling-based motion planners has been intensively studied in robotics, but they have been applied also in other research fields.
%In bio-chemistry, the planners have been used to study
to study loop motions~\cite{cortes2004geometric}
%protein folding~\cite{amato2002using,raveh2009rapid,novinskaya2015improving,songPFintro},
and for protein folding~\cite{raveh2009rapid,novinskaya2015improving,amato2002using}.
%or for tunnel detection~\cite{vonasek2017tunnel}.

@@

In the expansion step, each conformation from $\L$ is tested in order to approach $\qrand$.
By using more conformations, there is a higher chance that at least one of them can expand the tree towards the random sample $\qrand$.
On the other hand, using a large number of conformations also increases the time complexity of the expansion step.
Selection of a suitable set of conformations $\L$ is therefore crucial.
Depending on the ligand, some conformations can be considered as more packed i.e., the atoms in such conformations are closer to each other, but 
the atoms can also form a more elongated shape (Fig.~\ref{fig::m040c}).
From the motion planning point of view, both types of conformations are useful.
The longer conformations can better fit in a narrow passage, while the packed conformations can be better rotated at a given position.
In the narrow parts of the tunnel, the tree can be extended by the longer conformations, while in the wider regions, also the
packed conformations can be used.
\vspace{-3mm}

\begin{figure}[hb]
\centering
%\includegraphics[width=0.25\textwidth]{fig/m040-conformations}
\includegraphics[width=0.14\textwidth]{fig/m040-conf1} \hskip 25pt
\includegraphics[width=0.11\textwidth]{fig/m040-conf2}
\caption{\label{fig::m040c}
 The examples of a `long' conformation (left) and a `packed' one for the ligand m040 (1,5-Dichloropentane).
}
\end{figure}

@@

The results are summarized in Tab.~\ref{tab::main}.
The table shows the traversability rate only for the main tunnel, as no trajectory was found in the side tunnel for most of the tested ligands.

\begin{table}[bt]
\centering
\caption{\label{tab::main}
    Traversability of the tunnels for ligands with 50 conformations. 
    The number after '$/$' denotes the number of atoms.
}
\small
\renewcommand{\tabcolsep}{4.3pt}
{\small
\renewcommand{\arraystretch}{0.7}
\input graph.tex
}
\end{table}

The only exception is the ligand m037t, which was the smallest tested one (with 8 atoms), and therefore it could fit even inside the side tunnel.
The traversability rate is higher for the most scaled-down ligands ($\smin=0.5$) and it decreases with the increasing scale.
For the scale $\smin=0.8$, the most traversable frames were detected for ligands DCP (8~\%), m037t (20~\%) and m038t (16~\%).

@@

\vspace{-3mm}
The most time-consuming part of the planners is the collision detection. 
Although both methods can utilize the same number of collision detection queries in the expansion step, \RB\ is slower than \RA.
The reason is that \RA\ expands the tree more closer to the random samples $\qrand$ than \RB.
This results in a faster exploration of the configuration space and therefore, \RA\ needs less number of iterations to reach the active site.
The better performance of \RA\ is caused by the proposed atom-based metric, because this distance depends on the actual shape of the ligand and 
it allows the ligands to retract more towards $\qrand$.
On the contrary, \RB\ utilizes the single 6D Euclidean weighted metric in the expansion, where the weights are the same for all conformations.
Depending on these weights, the metric favors samples that differ from $\qrand$ more in the rotation or translation part.
However, the weights between the rotation/translation parts should depend on the shape of the ligand, which is different in each conformation.
By using a single set of weights in the 6D Euclidean metric, we implicitly favor either the packed or the longer conformations.
%To improve performance of \RB, the weights of the 6D Euclidean metric should differ for each conformation, but it would
%prevent the usage KD-tree for the nearest-neighbor search.
